The Magnetorotational Instability Prefers Three Dimensions

### 2019-07-22 paper/paper.tex
Lots of updates to text.

%\documentclass[aps,prl,preprint,groupedaddress]{revtex4-1}
%\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
\documentclass[aps,prl, reprint,superscriptaddress]{revtex4-1}
\newcommand{\SSC}{S/S_{c}}
\usepackage{graphicx}
% You should use BibTeX and apsrev.bst for references
% Choosing a journal automatically selects the correct APS
% BibTeX style file (bst file), so only uncomment the line
% below if necessary.
%\bibliographystyle{apsrev4-1}

@@

\begin{abstract}
Write abstract last.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}

\begin{itemize}
\item 3D $\to$ dynamo
\item stars, NSSL
\item experiments
\end{itemize}

\section{Method}
\label{sec:method}

We solve the linearized magnetohydrodynamic equations in a frame rotating with angular frequency $\Omega$. The Navier-Stokes equations is in a fairly standard form,
\begin{equation}
  \label{eq:ns}
  \frac{D \mathbf{v}}{Dt} - f \hat{z} \times \mathbf{v} + S v_x \hat{z} + \mathbf{\nabla}{p} + \nu \mathbf{\nabla} \times \mathbf{\omega} = 0,
\end{equation}
where $\omega = \mathbf{\nabla} \times \mathbf{v}$ is the vorticity, $f = 2 \Omega$ is the Coriolis parameter, and $S$ is the background shear rate. $Df/Dt = \partial f/\partial t + S x \partial f/\partial y$ is the advective derivative including the background linear shear velocity.  We write the induction equation in terms of the $x$ component of the magnetic field,
\begin{equation}
  \label{eq:Bx}
  \frac{D B_x}{Dt} - B_0 \partial_z v_x + \eta (\partial_y J_z - \partial_z J_y) = 0
\end{equation}
and the $x$ component of the current density,
\begin{equation}
  \label{eq:Jx}
  \frac{D J_x}{Dt} - B_0 \partial_z \omega_x + S \partial_z B_x - \eta \nabla^2 J_x = 0
\end{equation}
We explicitly enforce the divergence constraint on the incompressible velocity field $v$ and the magnetic field,
\begin{equation}
  \label{eq:divu}
  \mathbf{\nabla} \cdot \mathbf{v} = \mathbf{\nabla} \cdot \mathbf{B} = 0.
\end{equation}
The magnetic field is expressed in Alfven units, $\mathbf{B} = \mathbf{B_*}/\sqrt{\mu_0 \rho}$.
We consider the behavior of the MRI in a vertically ($z$) and horizontally ($y$) periodic channel of width $L_x = \pi$. The walls at $x = \pm L_x/2$ are perfectly conducting and stress-free. The MRI is a weak-field instability; in the invisicd, ideal case the critical shear rate for instability in two dimensions is given by
\begin{equation}
  \label{eq:Sc}
  S_c = \frac{-\pi^2 B^2}{f d^2}.
\end{equation}
Here, we use $\SSC$ as our control parameter.

We solve equations~(\ref{eq:ns})--~(\ref{eq:divu}) by assuming harmonic perturbations in $y$ and $z$, $f = \hat{f}(x) e^{i(k_y y + k_z z) + \gamma t}$. This reduces to an eigenvalue problem in $x$, which we solve using the \emph{Dedalus} pseudospectral framework. The advantage of this particular form is that it can be expressed as a set of ten first-order equations with Dirichlet boundary conditions for either no-slip or stress-free boundary conditions.

Our interest is in nearly ideal ($\eta = 0$), inviscid ($\nu = 0$) conditions; we thus set $\eta=\nu=10^{-5}$, though we have confirmed our main results are insensitive to this particular value.

\section{Results}
\label{sec:results}
Our most surprising result is that at very small values of supercriticality, the fastest growing mode is not axisymmetric. Figure~\ref{fig:growth_rate} shows the growth rates for three values of $\SSC$. At $\SSC = 1.01$, the maximum growth rate is $(k_y, k_z) \simeq (0.263, 0.447)$.

\begin{figure}[h]
  \includegraphics[width=\columnwidth]{run_11_output_growthrates.png}
  \includegraphics[width=\columnwidth]{run_12_output_growthrates.png}
  \includegraphics[width=\columnwidth]{run_13_output_growthrates.png}
  \caption{Growth rate as a function of $k_z$ and $k_y$. The maximum growth rate is far from $k_y = 0$, indicating the MRI is most unstable in a fully three-dimensional configuration.}
  \label{fig:growth_rate}
\end{figure}

Numerical simulations have consistently axisymmetric modes dominating early evolution of the MRI (cite Stone 95, etc) before parasitic modes cause a breakdown into MHD turbulence. However, our results recover this limit, as $\SSC$ becomes large. Figure~\ref{fig:alpha} plots $\alpha = \arctan k_y/k_z$ as  function of $\SSC$. Above $\SSC \gtrsim 2$, $\alpha$ is zero, indicating that axisymmetric modes have the fastest growth rates.

\begin{figure}[h]
  \includegraphics[width=\columnwidth]{alpha_vs_ssc_grid.pdf}
  \caption{Angle $\alpha$ vs $\SSC$. Stability limit is at $\SSC \simeq 0.102$}
  \label{fig:alpha}
\end{figure}
\section{Non-Normality}
\label{sec:non-normality}

@@

% Create the reference section using BibTeX:
\bibliography{basename of .bib file}

>>>>

%\documentclass[aps,prl,preprint,groupedaddress]{revtex4-1}
%\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
\documentclass[aps,prl,reprint,superscriptaddress]{revtex4-1}
\newcommand{\SSC}{S/S_{c}}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}
\newcommand{\apjs}{The Astrophysical Journal Supplement Series}
\usepackage{graphicx}
% You should use BibTeX and apsrev.bst for references
% Choosing a journal automatically selects the correct APS
% BibTeX style file (bst file), so only uncomment the line
% below if necessary.
%\bibliographystyle{apsrev4-1}

@@

\begin{abstract}
The magnetorotational instability (MRI) is a powerful instability in which conducting fluids with weak magnetic fields and decreasing angular velocity are linearly unstable.
The MRI is thought to be essential to the operation of astrophysical accretion disks, where the angular velocity graident is supplied by the gravitational field of the central object.
However, the MRI may also be important in the internal shear layers of stars, where the MRI may operate in a poorly studied regime where the shear is near critical but the dissipative forces are negligable.
Here, we consider the linear stability of three-dimensional perturbations to a nearly inviscid and ideal magnetofluid and show that the fastest growing modes are three dimensional when the instability is near onset.
These three dimensional modes are unstable at shear rates down to 0.10 $S_c$, the critical value in two dimensions, and they have growth rates exceeding the fastest two dimensional modes until the shear parameter is about $2.05 S_c$.
These modes are significant, as they suggest the possibility of direct dynamo action from the linear instability itself.
Furthermore, even when two dimensional modes have the fastest growth rates, significant numbers of three dimensional modes are unstable.
In such a system, the non-normal character of the MRI cannot be ignored anytime the shear is near critical.
Because evidence from weakly non-linear analysis suggests that the MRI saturates by pushing the shear rate to its critical value, non-normality is never unimportant for systems, like stars, that can rearrange their angular velocity profiles.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
The magnetorotational instability (MRI) is an extremely important instability in astrophysical fluid dynamics.
By changing the stabiliity criterion for differentially rotating flows from a negative angular \emph{momentum} gradient to a negative angular \emph{velocity} gradient, a weak magnetic field allows the Keplerian velocity profile due to a central point mass to drive turbulence \citep[e.g.][]{1998RvMP...70....1B}.
This discovery provided a pathway to explain the ubiquitous accretion onto compact objects at rates compatible with observations and may play a significant role in the formation of planets \citep[e.g.][]{2007Natur.448.1022J}.
In these objects, the gravitational field dominates the local plasma dynamics, and thus the MRI cannot significantly affect the background shear.
It must saturate by other means in these situations \citep{2018MNRAS.474.3451X}.
However, stars have differential rotation profiles driven by much weaker stresses.
In these regions, the MRI could be active and because the driving is much weaker, it can sature by pushing the background shear close to critical \citep{2017ApJ...841....1C,2017ApJ...841....2C}.

Here, we investigate the linear stability of three dimensional perturbations near the critical shear rate $S_c$ for a nearly invisicd, ideal MHD flow.
We find that the first modes to become destabilized are three-dimensional and thus act as a dynamo even in the absence of secondary instability.
These results strongly suggest that the non-normality of the linear operator leading to the MRI are always important, even near onset.
% \begin{itemize}
% \item 3D $\to$ dynamo
% \item stars, NSSL
% \item experiments
% \item Squire's theorem violation? Check MHD Squire, check S instead of Re...
% \item Do eigenvalue solution for B growth? i.e.\ check if it is a dynamo?
% \end{itemize}

\section{Method}
\label{sec:method}
We solve the linearized magnetohydrodynamic equations in a frame rotating with angular frequency $\Omega$. The Navier-Stokes equation is in a fairly standard form,
\begin{equation}
  \label{eq:ns}
  \frac{D \mathbf{v}}{Dt} - f \hat{z} \times \mathbf{v} + S v_x \hat{z} + \mathbf{\nabla}{p} + \nu \mathbf{\nabla} \times \mathbf{\omega} = 0,
\end{equation}
where $\omega = \mathbf{\nabla} \times \mathbf{v}$ is the vorticity, $f = 2 \Omega$ is the Coriolis parameter, and $S$ is the background shear rate. $Df/Dt = \partial f/\partial t + S x \partial f/\partial y$ is the advective derivative including the background linear shear velocity.  We write the induction equation in terms of the $x$ component of the magnetic field,
\begin{equation}
  \label{eq:Bx}
  \frac{D B_x}{Dt} - B_0 \partial_z v_x + \eta (\partial_y J_z - \partial_z J_y) = 0
\end{equation}
and the $x$ component of the current density,
\begin{equation}
  \label{eq:Jx}
  \frac{D J_x}{Dt} - B_0 \partial_z \omega_x + S \partial_z B_x - \eta \nabla^2 J_x = 0
\end{equation}
We explicitly enforce the divergence constraint on the incompressible velocity field $v$ and the magnetic field,
\begin{equation}
  \label{eq:divu}
  \mathbf{\nabla} \cdot \mathbf{v} = \mathbf{\nabla} \cdot \mathbf{B} = 0.
\end{equation}
The magnetic field is expressed in Alfven units, $\mathbf{B} = \mathbf{B_*}/\sqrt{\mu_0 \rho}$.
We consider the behavior of the MRI in a vertically ($z$) and horizontally ($y$) periodic channel of width $L_x = \pi$. The walls at $x = \pm L_x/2$ are perfectly conducting and stress-free. The MRI is a weak-field instability; in the inviscid, ideal case the critical shear rate for instability in two dimensions is given by
\begin{equation}
  \label{eq:Sc}
  S_c = \frac{-\pi^2 B^2}{f d^2}.
\end{equation}
Here, we use $\SSC$ as our control parameter.

We solve equations~(\ref{eq:ns})--~(\ref{eq:divu}) by assuming harmonic perturbations in $y$ and $z$, $f = \hat{f}(x) e^{i(k_y y + k_z z) + \gamma t}$. This reduces to an eigenvalue problem in $x$, which we solve using the \emph{Dedalus} framework. The advantage of this particular form is that it can be expressed as a set of ten first-order equations with Dirichlet boundary conditions for the stress-free, perfect conductor boundary conditions we use.

Our interest is in nearly ideal ($\eta = 0$), inviscid ($\nu = 0$) conditions.
However, in order to avoid critical layers in the stable solutions, we set $\eta=\nu=10^{-5}$.
We have confirmed our main results are insensitive to this particular value.

\section{Results}
\label{sec:results}
Our most surprising result is that at very small values of supercriticality, the fastest growing mode is not axisymmetric. 
Figure~\ref{fig:growth_rate}a shows the growth rates for four values of $\SSC$. 
At $\SSC = 1.02$, the maximum growth rate is $(k_y, k_z) \simeq (0.263, 0.447)$.
Figure~\ref{fig:growth_rate}b compares the numerically growth rates for $\SSC=1.002$ between $0 <= k_y < 0.2$ to an asymptotic calculation.
By expanding equations~(\ref{eq:ns}-~\ref{eq:divu}) in terms of $S = \SSC + \epsilon R$, we compute the leading order correction to the growth rate from three-dimensional effects as
\begin{equation}
  \label{eq:asymp}
  \left(\pi ^2 B^2+d^2 f^2 \right) \omega^{2} - B^4 k^2 \left(d^2 k^2-\pi ^2 R\right)   \ + \ \frac{\pi ^2 B^4 \ell ^2}{12}  \left(\left(6+\pi ^2\right) q+\pi ^2-6\right) \ = \ 0.
\end{equation}

%
\onecolumngrid

\begin{figure*}[b!]
  \includegraphics[width=\textwidth]{fig_1.pdf}
  \caption{Stuff}
  \label{fig:growth_rate}
\end{figure*}
\twocolumngrid
%
Numerical simulations have consistently shown axisymmetric modes dominating early evolution of the MRI before parasitic modes cause a breakdown into MHD turbulence \citep[e.g.][]{1995ApJ...440..742H,2018ApJ...853..174H,2019ApJS..241...26D}. 
In order to reconcile our results with these simulations, in figure~\ref{fig:alpha} we plot the angle $\alpha = \arctan k_y/k_z$ of the fastest growing mode as function of $\SSC$.
Above $\SSC \gtrsim 2.05$, $\alpha$ is zero, indicating that axisymmetric modes have the fastest growth rates.
For the fiducial run in \citet{1996ApJ...464..690H}, $\SSC \simeq 4.84$ (similar values are found in most works); thus we recover these results.

\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{alpha_vs_ssc_grid.pdf}
  \caption{Angle $\alpha$ vs $\SSC$. Stability limit is at $\SSC \simeq 0.102$}
  \label{fig:alpha}
\end{figure}

\section{Non-Normality}
\label{sec:non-normality}

@@

% Create the reference section using BibTeX:
\bibliography{mri}
